Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-27 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Completed Draft

\Title{Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard}
\Author{Daniel Madrzykowski}
\Year{2016}

@@

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs~\cite{McGrattan:2014}. More than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

@@

A statistician, George E.P. Box, is credited with the quote, `Essentially, all models are wrong, but some are useful.'  The ability of the numerical methods to simulate the results of the physical experiments was mixed.  Limited in some cases, by the available input data, inappropriate application, or lack of rigor in the underlying physics.  That said the models tended to provide enough inside to get a user `in the ballpark'.

@@

In Chapter 1, as the investigator, you had completed your review of the scene and had collected evidence which included an electric heater.  You are not sure sure about the fire's progression or the area of origin.  Before leaving the scene, you meet up with the fire captain and begin asking him about the fire.  The captain tells you that the doors and windows of the townhouse were closed upon arrival.  When the front door was opened heavy smoke flowed out, followed by flames.  Other fire fighters were venting the upstairs windows.  The fire grew and burned for a short time before water was applied to the   
flow path
inconsistent fire pattern near the right side of the sofa.  It appears that pattern was generated pre-flashover.  The pattern height is on the order of   calculate HRR and FH for cushion.
air equation

@@

Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires

The main objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  The first step was to examine the repeatability of different fire sources.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, $\pm$ 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The fire pattern heights generated on painted gypsum wall board by the natural gas burner and gasoline fueled fires were shown to have uncertainties of $\pm$ 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits. The patterns generated by the polyurethane foam fueled fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.    The comparison of the fire pattern heights and the mean flame height demonstrated that the “steady state” natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.  This trend was the same as exhibited by the repeatability of the heat release rate measurements.  Given that heat release rate is the source of the energy which was transferred to the walls in order generate the fire patterns, it stands to reason that the greater the uncertainty in the heat release rate the greater the uncertainty in pattern and any associated thermal measurements.

@@

Examining the capability of numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench-scale and full-scale fire experiments conducted in this study.

The second objective of this study was to examine the capability of numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench-scale and full-scale fire experiments conducted in this study.  The numerical tools used ranged from simple empirical algorithms to a computational fluid dynamics model.  Each of the tools are freely available to investigators and many of the tools have been packaged in a form that allows the user to input the ``data'' and generate a result.  There is software available to aid with the development of the computational fluid dynamics model.  The first challenge in using the numerical tools is finding the appropriate data to us as input.

Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires
Examining the capability of numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench-scale and full-scale fire experiments conducted in this study.

This manuscript highlights the many uncertainties involved in what appear to be simple fire experiments.  Uncertainties related to fuel type, measurement methods, and analysis techniques were examined.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability between replicates was greater for more complex fuels.  The variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%.  However, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.

The models are useful tools for gaining insight into fire behavior but in many cases many require specific experimental data for input or for validation. The availability   of well characterized input data is limited and data for newer materials is unavailable. Additional research is needed for modeling data development protocols.  Improved data repositories are also needed. Further research is needed to understand the appropriate resolution for implementing the results based on fuel type, fuel geometry, and ventilation.  Most importantly, there is a critical need to extend this research to better understand the capabilities and limitations of predictive methods using fuels and scenarios that fire investigators are likely to encounter during a fire investigation.

>>>>

\Title{Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard}
\Author{Daniel Madrzykowski}
\Year{2017}

@@

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs~\cite{FDS_Verification_Guide,FDS_Validation_Guide}. More than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

@@

A statistician, George E.P. Box, is credited with the quote, `Essentially, all models are wrong, but some are useful.'  The ability of the numerical methods to simulate the results of the physical experiments was mixed.  Limited in some cases, by the available input data, inappropriate application, or lack of rigor in the underlying physics.  That said the models tended to provide enough insight to get a user `in the ballpark'.

@@

In Chapter 1, as the investigator, you had completed your review of the scene and had collected evidence which included an electric heater.  You are not sure sure about the fire's progression or the area of origin.  Before leaving the scene, you meet up with the fire captain and begin asking him about the fire.  The captain tells you that the doors and windows of the townhouse were closed upon arrival.  When the front door was opened heavy smoke flowed out, followed by flames.  Other fire fighters were venting the upstairs windows.  The fire grew and burned for a short time before water was applied through the front door and the fire was knocked down.

Comparing the captain's description with your scene, it would seem that the fire started and burned as a fuel-limited fire, before transitioning to a ventilation limited 
fire, where the heat and fuel where adequate to sustain combustion, but perhaps oxygen was in short supply.  This fits the description of the fire flaring up and exiting the front door, shortly after it was opened.  With the information that other firefighters were opening the upstairs windows at this time, you starting thinking about the flow path in the town home.  The front door served as a bi-directional vent with the lower portion serving as an air inlet and the upper portion serving as an exhaust for flames.  In the upstairs bedroom there was mostly smoke damage and no significant heat damage. It seems that the open bedroom window served as a smoke exhaust vent.

The large V pattern on the wall behind the sofa and the stair wall may have been generated after the door and upstairs window was opened. This is an area where the fuel rich hot fire gases would mix with fresh air entering the front door and burn.  Area of the living room that was more remote from the front door, the source of oxygen, did not have as much fire damage.  You remember that there was fire damage to the wall behind the right side of the sofa.  This burned area on the wall seems distinct and disconnected from the area with the most fire damage, which is to the left of the sofa.  You decide to go back in the town house for another look, Figure~\ref{Pre_Flashover_Pattern}.  The pattern behind the sofa had to be generated early in the fire event as it is away from the front door. Further the wall behind the sofa between this pattern and the larger area of fire damage closer to the front door was protected and not exposed to flaming combustion.  The inside of the sofa's seat frame was charred on the inside but not on the outside in the area in front of this pattern.  The bottom of pattern is located near the base of the sofa seat cushion.  The pattern height is approximately 0.7~m high extending just above the top of the wood sofa frame.  The base of the pattern is approximately 0.3~m.

Using Heskestad's flame height equation and solving for the heat release rate based on the pattern height yields a heat release rate on the order of 55~kW. Unknowns include the impact of the back portion of the sofa, but this fire size seems reasonable to you with regard to the vision of a sofa cushion burning. This information coupled with sequence of events provided by the fire captain and other fire damage would indicate that the right side of the sofa is likely the area of origin, even though it does not have the most fire damage.  You still need to interview the occupants, was an accidental ignition due to careless smoking possible? or a lithium ion battery powered device?  Also you are still going to check if the heater switch was in the on position, if not that further supports your current hypothesis about the right side of the sofa.  You will also take samples of the debris in the area of the right seat cushion to send to the lab.  You have done all that you can at the scene.

If needed, you have the dimensions of the house and the opening so you could have a computer model of the house made to examine how long it would take a small fire to decrease the level of the oxygen within the house to 15~\%.  This might be needed to get insight on a time line that you might get from the interviews of the occupants.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/Pre_Flashover_Pattern} \\
	\caption[Photograph showing the wall with the stand alone fire pattern toward the right side of the sofa.]{Photograph showing the wall with the stand alone fire pattern toward the right side of the sofa.}
	\label{Pre_Flashover_Pattern}
\end{figure}

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/Table} \\
	\caption[Photograph showing the charred inside of the sofa's seat frame.]{Photograph showing the charred inside of the sofa's seat frame.}
	\label{Table}
\end{figure}

N.B. This section was included to provide a sense of how a fire investigator might use their knowledge of fire behavior, fire patterns, and numerical methods to assist them in following the scientific method.  Fire investigation requires significantly more than pattern interpretation, the totality of facts from the whole situation must gathered to the extent possible and then after analysis of the evidence, a fire cause may be determined.

@@

The main objective of this research study was to examine the repeatability of pre-flashover fire patterns on gypsum wallboard.  The patterns were generated from exposure to short duration (300 s maximum) well characterized fires.

The first step was to examine the repeatability of different fire sources.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel source had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  The variation in the average peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, $\pm$ 11~\%.  However the variation in the average peak heat release rates of the gasoline and the polyurethane foam increased due to uncertainties in the burning behavior of the fuels.  The average peak heat release rate of gasoline had a 95~\% confidence limit of $\pm$30~\% and the same value for the polyurethane foam was $\pm$47~\%.

The fire pattern heights generated on painted gypsum wallboard by the natural gas burner and gasoline fueled fires were shown to have uncertainties of $\pm$ 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits. The patterns generated by the polyurethane foam fueled fire, had more than twice the variability of the natural gas and the gasoline fire patterns.  This was likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.

The comparison of the fire pattern heights and the mean flame heights demonstrated that the “steady state” natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.  This trend was the same as exhibited by the repeatability of the heat release rate measurements.  Given that the heat release rate from the fires was the source of the energy transferred to the walls to generate the fire patterns.  It stands to reason that the greater the uncertainty in the heat release rate, the greater the uncertainty in the resulting pattern and any associated thermal measurements.

@@

Another objective of this study was to examine the capability of numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench-scale and full-scale fire experiments conducted as part of this study.  The numerical tools used ranged from simple empirical algorithms to a computational fluid dynamics model.  Each of the tools are freely available to investigators and many of the tools have been packaged in a form that allows the user to input the ``data'' and generate a result.  There is also graphical user interface software available to aid with the development of the computational fluid dynamics model.  The first challenge in using the numerical tools is finding the appropriate data to use as input.

A range of methods used to simulate characteristics of fires were applied using data from the source fire experiments as input.  The algebra based predictions for heat flux to a target under predicted the measured values.  The gasoline fueled fires had the best agreement of 22~\% or less for the heat flux predictions.  The polyurethane foam results exhibited the worst agreement with difference of at least 40~\% between the measured and calculated values.  Improved understanding of the radiative fraction from different fuels would improve the capability of the predictions.

The best agreement between measured flame height, the measured height of the fire pattern, and flame height predictions from Heskestad's formula occurred with the gasoline fueled fires.  Given the overlap of the expanded uncertainty and predicted range of values, the measured and predicted heights would be considered similar.  The gasoline fueled fires released the highest amount of energy and had a higher radiative fraction than the natural gas and polyurethane foam fueled fires. The flame height algorithm tended to over predict the values for the natural gas and the polyurethane fueled fires.

The agreement between the values predicted by the computational fluid dynamic model, FDS, and the measurements from these experiments demonstrated examples of agreement that were within the measurement uncertainty estimates as well as examples with poor agreement.  The differences were a result, in some cases, by limitations within the combustion sub-model close to the burner surface.  In those cases as the distance between the burner surface increased the predicted values tended to converge with the measured values.

The wall located adjacent to the burner had a significant impact on the flow field in and around the flame and plume region.  The horizontal flame movements observed during the experiments were not simulated to the same degree by the model.  Since areas around the flame have steep thermal gradients small differences in position can result in large differences in temperature and heat flux.

The models proved to be useful tools for gaining insight into fire behavior but in many cases the models require specific experimental data for input or for validation. The availability of well characterized input data is limited and data for newer materials is unavailable. Additional research is needed for modeling data development protocols.  Improved data repositories are also needed. Further research is needed to understand the appropriate resolution for implementing the results based on fuel type, fuel geometry, and ventilation.  Most importantly, there is a critical need to extend this research to better understand the capabilities and limitations of predictive methods using fuels and scenarios that fire investigators are likely to encounter during a fire investigation.
